---
title: "Group2.Rmd"
author: "Tamar Yastrab & Hasanat Jahan"
date: "6/23/2020"
output: html_document
---

We reproduced the findings in Predicting the Present with Google Trends with both author provided data and data we gathered independently
<http://people.ischool.berkeley.edu/~hal/Papers/2011/ptp.pdf>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r packages}
library(tidyverse)
library(scales)
library(modelr)
library(lubridate)
library(dplyr)
```

# Reproduce the findings based on author-provided data

Read in the given data
```{r Downloading the data}
mvp <- read.csv("merged.csv", header=T) %>% mutate(last_month = lag(sales), last_year = lag(sales, 12)) %>% mutate(row_num = row_number()) %>% mutate(log_sales = log(sales)) %>% mutate(ymd = as.Date(Period, "%Y/%m/%d"))
```

We created 2 models for the mvp data. The first only considers the previous month and the same month one year prior, and the second considers additional factors (insurance, suvs) provided by the trends data. 
```{r mvp model}

mvp_base_model <- lm(sales ~ last_month + last_year,  data=mvp)
summary(mvp_base_model)

mvp_trends_model <- lm(sales ~ last_month + last_year + suvs + insurance,  data=mvp)
summary(mvp_trends_model)

```

We created 2 vectors to keep track of the predictions made by the two models. We loop from 2005-06-01 to 2011-07-01 as indicated by the paper. For each iteration, we make a linear model that will serve as the base for a prediction and then add the prediction to the corresponding vector. 
```{r mvp predictions}
base_prediction = c()
trends_prediction = c()
mae_base_prediction = c()
mae_trends_prediction = c()

K <- 17: nrow(mvp) # dates

for (k in K) {
  test_data<- mvp %>% filter(row_num <= k) # only uses previous data
  model_base <- lm(log_sales ~ last_month + last_year , data=test_data)
  model_trends <- lm(log_sales ~ last_month + last_year + suvs + insurance, data=test_data)
  
  base_prediction[k] <- predict(model_base, test_data)[k]
  trends_prediction[k] <- predict(model_trends, test_data)[k]

  # calculate mae for trends and baseline
  mae_base_prediction[k] <- mean(abs(predict(model_base, test_data)[k] - test_data$log_sales))
# calculate mae for trends and baseline
  mae_trends_prediction[k] <- mean(abs(predict(model_trends, test_data)[k] - test_data$log_sales))

}
```

```{r calculating error}
mae_data <- data.frame(mae_base_prediction, mae_trends_prediction) %>%   
  na.omit(mae_base_prediction) %>% mutate(row_num = row_number() + 16)

mae_total <- mvp %>% left_join(mae_data, by = "row_num")

colors <- c("mae_trends_prediction" = "blue", "mae_base_prediction" = "black")

ggplot(mae_total, aes(x = ymd, y = mae_base_prediction)) +
  geom_line() +
  geom_line(aes(y = mae_trends_prediction, color = "mae_trends_prediction"), alpha= 0.4) +
  geom_line(aes(y = mae_base_prediction, color="mae_base_prediction")) +
  labs(x= 'Index', y ='MAE',color = "Legend") +
  scale_color_manual(values=colors) +
  scale_y_continuous()
```


```{r mvp ggplot}

prediction_data <- data.frame(base_prediction, trends_prediction) %>% na.omit(base_prediction) %>% mutate(row_num = row_number() + 16)

total <- mvp %>% left_join(prediction_data, by = "row_num")

pred_colors = c("Trends Prediction" = "red", "Base Prediction" = "grey", "log(mvp)"= "black")
ggplot(total, aes(x = ymd, y = log_sales)) +
  geom_line(aes(color = "log(mvp)")) +
  geom_line(aes(y = trends_prediction, color = "Trends Prediction"), linetype = "dotdash") +
  geom_line(aes(y = base_prediction, color="Base Prediction")) +
  labs(x='Index',
       y='log(mvp)', 
       color="Legend") +
  scale_color_manual(values = pred_colors) +
  scale_y_continuous()

```

# Getting the Data from their original locations

Rather than just take the author-provided data, we wanted to get the data ourselves from the Census Bureau and Google Trends. We cleaned the data and joined the mvp data together with the trends data just like we did above. The rest of the code is identical but with the new data. 

```{r real data from the internet}
#real_base <- read_tsv("base_mvp.tsv") %>% filter(JAN >= 60000)
real_trends_mvp <- left_join(read.csv("real_trends_insurance.csv", header = T), 
                         read.csv("real_trends_trucks.csv", header = T), 
                         by = "Month") %>% 
  rename("insurance" = Geo..United.States.x, "suvs" = Geo..United.States.y) %>% 
  mutate(ymd2 = paste(Month, "-01", sep="")) %>% 
  mutate(ymd = as.Date(ymd2, "%Y-%m-%d")) %>% 
  select(ymd, insurance, suvs)

real_base_mvp <- read.csv("real_base_autos.csv") %>% 
  mutate(ymd2 = paste(Period, "-01", sep="")) %>% 
  mutate(ymd = as.Date(ymd2, "%b-%y-%d")) %>% select(ymd, Value)

real_mvp <- left_join(real_trends_mvp, real_base_mvp, by = "ymd") %>% mutate(last_month = lag(Value), last_year = lag(Value, 12)) %>% mutate(row_num = row_number()) %>% mutate(log_sales = log(Value))

```


```{r real mvp model}

real_base_model <- lm(Value ~ last_month + last_year,  data=real_mvp)
summary(real_base_model)

real_trends_model <- lm(Value ~ last_month + last_year + suvs + insurance,  data=real_mvp)
summary(real_trends_model)

```


```{r real mvp predictions}
real_base_prediction = c()
real_trends_prediction = c()
real_mae_base_prediction = c()
real_mae_trends_prediction = c()

K <- 17: nrow(real_mvp) # dates

for (k in K) {
  test_data<- real_mvp %>% filter(row_num <= k) # only uses previous data
  model_base <- lm(log_sales ~ last_month + last_year , data=test_data)
  model_trends <- lm(log_sales ~ last_month + last_year + suvs + insurance, data=test_data)
  
  real_base_prediction[k] <- predict(model_base, test_data)[k]
  real_trends_prediction[k] <- predict(model_trends, test_data)[k]

  real_mae_base_prediction[k] <- mean(abs(predict(model_base, test_data)[k] - test_data$log_sales))
  real_mae_trends_prediction[k] <- mean(abs(predict(model_trends, test_data)[k] - test_data$log_sales))

}
```

```{r real calculating error}
real_mae_data <- data.frame(real_mae_base_prediction, real_mae_trends_prediction) %>%   
  na.omit(real_mae_base_prediction) %>% mutate(row_num = row_number() + 16)

real_mae_total <- real_mvp %>% left_join(real_mae_data, by = "row_num")

colors <- c("real_mae_trends_prediction" = "blue", "real_mae_base_prediction" = "black")

ggplot(real_mae_total, aes(x = ymd, y = real_mae_base_prediction)) +
  geom_line() +
  geom_line(aes(y = real_mae_trends_prediction, color = "real_mae_trends_prediction"), alpha= 0.4) +
  geom_line(aes(y = real_mae_base_prediction, color="real_mae_base_prediction")) +
  labs(x= 'Index', y ='MAE',color = "Legend") +
  scale_color_manual(values=colors) +
  scale_y_continuous()
```


```{r real mvp ggplot}

real_prediction_data <- data.frame(real_base_prediction, real_trends_prediction) %>% na.omit(real_base_prediction) %>% mutate(row_num = row_number() + 16)

real_total <- real_mvp %>% left_join(real_prediction_data, by = "row_num")

pred_colors = c("Real Trends Prediction" = "red", "Real Base Prediction" = "grey", "log(mvp)"= "black")
ggplot(total, aes(x = ymd, y = log_sales)) +
  geom_line(aes(color = "log(mvp)")) +
  geom_line(aes(y = real_trends_prediction, color = "Real Trends Prediction"), linetype = "dotdash") +
  geom_line(aes(y = real_base_prediction, color="Real Base Prediction")) +
  labs(x='Index',
       y='log(mvp)', 
       color="Legend") +
  scale_color_manual(values = pred_colors) +
  scale_y_continuous()

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
